---
title: "Bulk RNAseq report"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Gianluca Alessio Mariani"

params:
  input_samples_table: "WRITE_INPUT_PATH_HERE"
  input_main_template: "./txt_templates/_template_main.txt"
  input_singlecomp_template: "./txt_templates/_template_singlecomp.txt"
  output_folder_rmd: "WRITE_OUTPUT_PATH_HERE"
---

```{r}
knitr::opts_chunk$set(echo       = FALSE,
                      message    = FALSE,
                      warning    = FALSE,
                      cache      = FALSE,
                      autodep    = TRUE,
                      fig.align  = 'center',
                      fig.width  = 10,
                      fig.height = 8)
```

```{r load_lib}
library(dplyr)
library(rlang)
library(glue)
library(knitr)
```

```{r function}
gen_comparison <- function(samples_df,
                           comp_id,
                           contrast_variable,
                           other_design_variables = NULL,
                           e_filter,
                           c_filter,
                           e_descr = NULL,
                           c_descr = NULL,
                           exclude = NULL) {
  
  comp_descr <- paste(e_descr, "vs", c_descr)
  e_descr <- paste(e_descr, "(exp)")
  c_descr <- paste(c_descr, "(contr)")

  if (!(contrast_variable %in% names(e_filter))) {
    stop(paste("contrast_variable", contrast_variable, "not found in e_filter"))
  }
  if (!(contrast_variable %in% names(c_filter))) {
    stop(paste("contrast_variable", contrast_variable, "not found in c_filter"))
  }

  e_contrast_value <- e_filter[[contrast_variable]]
  c_contrast_value <- c_filter[[contrast_variable]]

  if (length(e_contrast_value) != 1 || length(c_contrast_value) != 1) {
    stop("contrast_variable must have only one values both in e_filter and in c_filter")
  }

  e_samples <- samples_df
  for (filter_name in names(e_filter)) {
    e_samples <- e_samples %>% filter(.data[[filter_name]] %in% e_filter[[filter_name]])
  }

  c_samples <- samples_df
  for (filter_name in names(c_filter)) {
    c_samples <- c_samples %>% filter(.data[[filter_name]] %in% c_filter[[filter_name]])
  }

  comp_df <- bind_rows(
    e_samples %>% mutate(comparison_ID = comp_id, sample_type = "e"),
    c_samples %>% mutate(comparison_ID = comp_id, sample_type = "c")
  ) %>%
    dplyr::select(comparison_ID, sample_type, sample)

  if (!is.null(exclude)) {
    comp_df <- comp_df %>% filter(!(sample %in% exclude))
  }

  return(list(
    comparison_df = comp_df,
    comp_id = comp_id,
    comp_descr = comp_descr,
    contrast_variable = contrast_variable,
    e_contrast_value = e_contrast_value,
    c_contrast_value = c_contrast_value,
    other_design_variables = other_design_variables,
    e_descr = e_descr,
    c_descr = c_descr
  ))
}
```

```{r init}
sample_info <- read.csv(params$input_samples_table, stringsAsFactors = FALSE)
main_rmd <- readLines(params$input_main_template)
singlecomp_rmd <- readLines(params$input_singlecomp_template)
```

# Instructions

Add to chunk_code_1 the list of comparisons you want to generate by copying and pasting the template below
Modify:
- contrast_variable: the variable to use for contrast (e.g. condition, time, etc), it must be present as a column name in the sample_info dataframe
- other_design_variables: other variables to include in the design (can be NULL), used to correct for batch effects or other covariates
- e_filter: a named list of filters to select experimental samples, the names and values must match column names and column values in sample_info dataframe and must include the names specified in contrast_variable and other_design_variables
- c_filter: same as e_filter but for control samples
- e_descr: description for experimental samples, used in report texts (choose freely, it is for your personal use)
- c_descr: same as e_descr but for control samples
- exclude: optional, a vector of sample names to exclude from the comparison (can be NULL)
- report_filename: the name of the Rmd report file to generate for this comparison

(template to copy and paste, ignore the triple quotes):
"""
comp_id <- comp_id + 1
comp_list[[length(comp_list) + 1]] <- gen_comparison(
  samples_df = sample_info,
  comp_id = comp_id,
  exclude = NULL,
  
  contrast_variable = 'condition_single',
  other_design_variables = NULL,

  e_filter = list(condition_single = 'shIFI6_1_Day1'),
  c_filter = list(condition_single = 'shSCR_Day1'),
  
  e_descr = 'shIFI6_1 Day1',
  c_descr = 'shSCR Day1',
)
comp_list[[length(comp_list)]]$report_filename <- '02_de_ifi6_sh1_day1.Rmd'
"""

```{r generate_comparisons_text, results='asis', echo = FALSE}

chunk_code_1 <- "
comp_id <- 0
comp_list <- list()

############################################################################### shIFI6_1_vs_scramble_at_day_X

comp_id <- comp_id + 1
comp_list[[length(comp_list) + 1]] <- gen_comparison(
  samples_df = sample_info,
  comp_id = comp_id,
  
  contrast_variable = 'condition_single',
  e_filter = list(condition_single = 'shIFI6_1_Day1'),
  c_filter = list(condition_single = 'shSCR_Day1'),
  
  e_descr = 'shIFI6_1 Day1',
  c_descr = 'shSCR Day1',
)
comp_list[[length(comp_list)]]$report_filename <- '02_de_ifi6_sh1_day1.Rmd'
"

chunk_code_2 <- "
##### Generate_list_of_comparisons

text_deseq_groups <- ''

if (length(comp_list) > 0) {
  for (comp in comp_list) {
    
    comp_df <- comp$comparison_df
    
    # Sample names
    exp_samples <- comp_df %>%
      filter(sample_type == 'e') %>%
      pull(sample)
    
    ctrl_samples <- comp_df %>%
      filter(sample_type == 'c') %>%
      pull(sample)
    
    # Descriptions
    comp_id <- comp$comp_id
    comp_descr <- comp$comp_descr
    e_descr <- comp$e_descr
    c_descr <- comp$c_descr
    
    # Comparison text
    comparison_text <- paste0('(', e_descr, ' vs ', c_descr, ')')
    
    text_deseq_groups <- paste0(text_deseq_groups,
      '\\n',
      '➤ Comparison ', comp_id, ', ', comp_descr, '\\n',
      '\\n\\t- **\"Treatment\" samples** (', e_descr, ', e):\\n',
      paste0('\\t\\t', exp_samples, collapse = '\\n'), '\\n',
      '\\n\\t- **Control samples** (', c_descr, ', c):\\n',
      paste0('\\t\\t', ctrl_samples, collapse = '\\n'), '\\n\\n'
    )
  }
} else {
  text_deseq_groups <- '⚠️ No valid group defined for DGE analysis.\\n'
}

text_html <- gsub('\\n', '<br>', text_deseq_groups)
text_html <- gsub('\\t', '&nbsp;&nbsp;&nbsp;&nbsp;', text_html)
cat(text_html)
"
```

```{r generate_and_run_final_chunk}
final_chunk <- (paste0("```{r generate_comparisons, results='asis', echo = FALSE}\n", chunk_code_1, chunk_code_2, "\n```\n"))

eval(parse(text = chunk_code_1))
```

```{r main_yaml}
main_yaml <- glue('
---
title: "Bulk RNAseq report"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
date: "`r format(Sys.time(), \'%d %B, %Y\')`"
author: "Gianluca Alessio Mariani"

params:
  sample_info_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/data/samples_list.csv"
  utils_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/src/__utils_rna_seq_functions.R"
  dds_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/data/deseq2.dds.RData"
  ribosomal_genes_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/src/genes2filter_symbol.csv"
  min_sample: 10
  logfc: 1
  qvalue: 0.05
  genelist1: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/data/UNIQUELIST_shIFI6_vs_shIFI6.csv"
  genelist2: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/data/UNIQUELIST_shIFI6_vs_shSCR.csv"
  gene_name_type: "SYMBOL"
---
')
```

```{r main_rmd}
final_chunk_lines <- strsplit(final_chunk, "\n")[[1]]
pos <- grep("{__sub__}", main_rmd, fixed = TRUE)

main_rmd <- c(
  main_rmd[1:(pos-1)],
  final_chunk_lines,
  main_rmd[(pos+1):length(main_rmd)]
)

final_rmd <- c(main_yaml, "", main_rmd)

writeLines(final_rmd, file.path(params$output_folder_rmd, "01_baseline_analysis.Rmd"))
```

```{r singlecomp_yaml}
singlecomp_yaml_list <- list()

for (i in seq_along(comp_list)) {
  df <- comp_list[[i]]$comparison_df
  exp_samples <- df$sample[df$sample_type == "e"]
  ctrl_samples <- df$sample[df$sample_type == "c"]
  e_descr <- comp_list[[i]]$e_descr
  c_descr <- comp_list[[i]]$c_descr
  comp_descr <- comp_list[[i]]$comp_descr
  contrast_variable <- comp_list[[i]]$contrast_variable
  e_contrast_value <- comp_list[[i]]$e_contrast_value
  c_contrast_value <- comp_list[[i]]$c_contrast_value
  design_vars <- c(contrast_variable, comp_list[[i]]$other_design_variables)

  # Converti i sample in YAML-friendly lista
  exp_yaml <- paste0("        - \"", exp_samples, "\"", collapse = "\n")
  ctrl_yaml <- paste0("        - \"", ctrl_samples, "\"", collapse = "\n")
  e_descr_yaml <- paste0("      e_descr: \"", e_descr, "\"")
  c_descr_yaml <- paste0("      c_descr: \"", c_descr, "\"")
  comp_descr_yaml <- paste0("  comp_descr: \"", comp_descr, "\"")
  design_vars_yaml <- paste0("    - \"", design_vars, "\"", collapse = "\n")
  contrast_variable_yaml <- paste0("  contrast_variable: \"", contrast_variable, "\"")
  e_contrast_value_yaml <- paste0("  e_contrast_value: \"", e_contrast_value, "\"")
  c_contrast_value_yaml <- paste0("  c_contrast_value: \"", c_contrast_value, "\"")
  
singlecomp_yaml <- glue('
---
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
date: "`r format(Sys.time(), \'%d %B, %Y\')`"
author: "Gianluca Alessio Mariani"

params:
  design_vars: 
{design_vars_yaml}
{contrast_variable_yaml}
{e_contrast_value_yaml}
{c_contrast_value_yaml}
  num_samples: 23
  num_replicates: 2
  batch_effect: null
  logfc: 1
  gsea_terms: 15
  qvalue: 0.05
  gsea_pval2filter_GO: 0.05
  gsea_pval2filter_HALLMARK: 0.05
  path_option: ""
  top_n_genes_used4ora: 1000
  gene_name_type: "symbol"
  sample_info_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/data/sample_list_variables.csv"
  utils_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/src/__utils_rna_seq_functions.R"
  dds_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/data/deseq2.dds.RData" 
  ribosomal_genes_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/src/genes2filter_symbol.csv"
  hallmark_file: "/hpcnfs/scratch/DIMA/gmariani/PROJECTS/rna/reports/projects/frascolla/frascolla_chemoresistance/src/h.all.v2025.1.Hs.symbols.gmt"
{comp_descr_yaml}
  exp:
    value:
{e_descr_yaml}
      exp_samples:
{exp_yaml}
  contr:
    value:
{c_descr_yaml}
      ctrl_samples:
{ctrl_yaml}
---
')
  
  singlecomp_yaml_list[[i]] <- singlecomp_yaml
}
```

```{r singlecomp_rmd}
for (i in seq_along(comp_list)) {
 final_singlecomp_rmd <- c(singlecomp_yaml_list[[i]], "", singlecomp_rmd)

 writeLines(final_singlecomp_rmd, file.path(params$output_folder_rmd, comp_list[[i]]$report_filename)) 
}
```
